---
title: "VTPEH 6270 - Check Point 01"
subtitle: "Basic R Skills"
author: "Fina Zou"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Introduction

The main question we want to address is: ***how has vaccination coverage changed through the years?***

# Loading Data

```{r, results='hide'}
# Load file
data_vaccination=read.csv("CDC_vaccination_pregnancy.csv")
# Preview the data
head(data_vaccination)
```

The data set is not labeled using best practices (e.g., uses unusual characters like "..."). We can clean the formatting using the `janitor` package.

```{r, results='hide'}
# install.packages("janitor")
# library(janitor)
library(janitor)
data_vaccination=clean_names(data_vaccination)
# Preview the cleaned data
head(data_vaccination)

```

# Eploring the Data


```{r, results='hide'}

unique(data_vaccination$vaccine)

```

The considered vaccines are Influenza and Tdap, which correspond to influenza and tetanus/diphtheria/pertussis respectively.

```{r, results='hide'}
unique(data_vaccination$dimension_type)

```

The considered individual characteristics are Age and Race and Ethnicity.


```{r, results='hide'}
unique(data_vaccination$dimension[data_vaccination$dimension_type=="Race and Ethnicity"])

```

The represented races and ethnicities are White, Non-Hispanic; Black, Non-Hispanic; Hispanic; Other or Multiple Races, Non-Hispanic.


```{r, results='hide'}
min(data_vaccination$survey_year_influenza_season, na.rm = TRUE)
max(data_vaccination$survey_year_influenza_season, na.rm = TRUE)

```

The study spans from 2012 to 2022.

```{r, results='hide'}

# Check class
class(data_vaccination$estimate)

# Check values to ensure class conversion will not erase any important data
unique(data_vaccination$estimate)

# Class conversion
data_vaccination$estimate <- as.numeric(data_vaccination$estimate)

# Check output
class(data_vaccination$estimate)

```

Vaccination coverage should be recorded as numeric. In the original dataset it is stored as character

# Focus Description of 18 years old and older of NY State

*Subset the data to represent only 18+ years old of NY State*

```{r, results='hide'}
data_ny_18plus <- data_vaccination[
  data_vaccination$geography=="New York" &
    data_vaccination$dimension_type=="Age" &
    data_vaccination$dimension==">=18 Years",
]
head(data_ny_18plus)
```


```{r, results='hide'}
library(ggplot2)
ggplot(data_ny_18plus, aes(x=survey_year_influenza_season,y=estimate,color=vaccine)) +
  geom_line(linewidth=1) + geom_point(size=2) +
  labs(title="Estimated Vaccination Coverage among Pregnant Women \nOver Time in New York State, Age >= 18 Years", x="Survey Year / Influenza Season", y="Estimated Vaccination Coverage (%)",color="Vaccine") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11,hjust=0.5),legend.position="bottom")

```

This plot shows trends in estimated vaccination coverage over time for Influenza and Tdap among adults aged 18 years and older in New York State.
The influenza vaccination coverage is relatively stable. The Tdap vaccination coverage rose rapidly in earlier years and remained higher and more stable across the study period.


```{r, results='hide'}

max(data_ny_18plus$estimate[data_ny_18plus$vaccine=="Influenza"],na.rm = TRUE)
```

Influenza vaccination peaked at 64.3%.


```{r, results='hide'}
mean(data_ny_18plus$estimate[data_ny_18plus$vaccine=="Influenza"],na.rm = TRUE)

```

Average Influenza vaccination coverage over the study period is 57.5%.

# AI Use Disclosure Statement

This document was completed with assistance from ChatGPT. AI was used to help recall R functions and to provide explanations and examples of their usage. All code and interpretations were reviewed and finalized by the author.

